---
title: "James RBIF 125 Final"
author: "Author: James Hinojosa"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: yes
    code_folding: hide
    number_sections: true
  pdf_document:
    toc: true
vignette: >
    %\VignetteIndexEntry{Final}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
fontsize: 15pt
editor_options: 
  chunk_output_type: console
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
body {
text-align: justify}
</style>


```{r, include = FALSE}

# Sets home directory 
homedir <- "C:/Users/ryan1/Desktop/Brandeis University/Courses/RBIF-125 Bioinformatics Software Engineering and AI ML/Week 8/Final_Project/" 

```


# Download data set from GEO 
The code below will define a function to download a data set from the Gene Expression Omnibus (GEO). This code provides a way to download either an Array or RNAseq dataset. 

```{r 1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Installs necessary packages if not already installed
# BiocManager::install("DESeq2")
# BiocManager::install("limma")
# install.packages("data.table")
# BiocManager::install("sva")
# install.packages('R.utils')
library(DESeq2); library(limma); library(data.table); library(ggplot2); library(sva); library(R.utils)

# Creates function to download data from GEO
GEODataDownload <- function(DS, gpl, gsm, PlateAnnotInfo, GenerateMetaData, Technology){
  library(DESeq2); library(limma); library(data.table)
  
  # Defines downloading an "Array" dataset
  if(Technology == "Array"){
    gset <- getGEO(DS)
    if(length(gset) > 1) idx <- grep(gpl, attr(gset, "names")) else idx <- 1
    gset <- gset[[idx]]
    fvarLabels(gset) <- make.names(fvarLabels(gset))
    comp <- gsub(" ", "", gsm)
    comp <- gsub(",", "", comp)
    gsms <- paste0(comp)
    
    # Sets up raw names
    sml <- c()
    for(i in 1:nchar(gsms)){ sml[i] <- substr(gsms,i,i)}
    ex <- exprs(gset)
    qx <- as.numeric(quantile(ex, c(0., 0.25, 0.5, 0.75, 0.99, 1.0), na.rm=T))
    LogC <- (qx[5] > 100) ||
      (qx[6]-qx[1] > 50 && qx[2] > 0) ||
      (qx[2] > 0 && qx[2] < 1 && qx[4] < 2)
    if(LogC){ ex[which(ex <= 0)] <- NaN
    exprs(gset) <- log2(ex) }
    sml <- paste("G", sml, sep="")
    f1 <- as.factor(sml)
    gset$description2 <- f1
    design <- model.matrix(~description2 + 0, gset)
    colnames(design) <- levels(f1)
    fit <- lmFit(gset, design)
    cont.matrix <- makeContrasts(G1-G0, levels = design)
    fit2 <- contrasts.fit(fit, cont.matrix)
    fit2 <- eBayes(fit2, 0.01)
    tT <- topTable(fit2, adjust="fdr", sort.by = "B", number = 25000000000)
    
    # Subsets data
    ex2 <- data.table(subset(tT, select=c("ID", "logFC", "P.Value", "adj.P.Val")))
    ex2$ID <- as.character(ex2$ID)
    
    # Annotates with gene names
    plat <- PlateAnnotInfo[GPLID == gpl,][,!"GPLID", with = FALSE]
    if(nrow(plat) == 0){ print(paste("There is no annotation information available for", gpl)) }
    plat$ID <- as.character(plat$ID)
    plat <- plat[!duplicated(plat$ID),]
    ex2 <- merge(plat, ex2, by = "ID")
    ex2$ID <- as.character(ex2$ID)
    exraw <- data.table(ex)
    exraw$ID <- as.character(rownames(ex))
    
    # Annotates raw data with gene names
    ex2 <- merge(ex2, exraw, by = "ID")
    
    # Generates Metadata
    Pdat <- pData(gset)
    
    # Adds Metadata
    Pdat <- as.data.table(Pdat)
    return(list(Data = ex2, MetaData = Pdat))
  }
  
  # Defines downloading an "RNAseq" dataset
  if(Technology == "RNAseq"){
    ACC <- paste("acc=", DS, sep = "")
    file <- paste("file=", DS, "_raw_counts_GRCh38.p13_NCBI.tsv.gz", sep = "")
    comp <- gsub(" ", "", gsm)
    comp <- gsub(",", "", comp)
    gsms <- paste0(comp)
    
    # Sets up DEG names
    urld <- "https://www.ncbi.nlm.nih.gov/geo/download/?format=file&type=rnaseq_counts"
    path <- paste(urld, ACC, file, sep="&");
    tbl <- as.matrix(data.table::fread(path, header=T, colClasses="integer"), rownames="GeneID")
    exraw <- tbl 
    apath <- paste(urld, "type=rnaseq_counts", "file=Human.GRCh38.p13.annot.tsv.gz", sep="&")
    annot <- data.table::fread(apath, header=T, quote="", stringsAsFactors=F, data.table=F)
    rownames(annot) <- annot$GeneID
    sml <- strsplit(gsms, split="")[[1]]
    sel <- which(sml != "X")
    sml <- sml[sel]
    tbl <- tbl[ ,sel]
    gs <- factor(sml)
    groups <- make.names(c("Ctrl", "Tx"))
    levels(gs) <- groups
    sample_info <- data.frame(Group = gs, row.names = colnames(tbl))
    keep <- rowSums( tbl >= 10 ) >= min(table(gs))
    tbl <- tbl[keep, ]
    ds <- DESeqDataSetFromMatrix(countData=tbl, colData=sample_info, design= ~Group)
    ds <- DESeq(ds, test="Wald", sfType="poscount")
    r <- results(ds, contrast=c("Group", groups[2], groups[1]), alpha=0.05, pAdjustMethod ="fdr")
    tT <- r[order(r$padj)[1:length(r$padj)],]
    tT <- merge(as.data.frame(tT), annot, by=0, sort=F)
    tT <- subset(tT, select=c("GeneID","padj","pvalue","lfcSE","stat","log2FoldChange","baseMean","Symbol","Description"))
    
    # Subsets data
    ex2 <- data.table(subset(tT, select=c("GeneID", "Symbol", "Description", "log2FoldChange", "pvalue", "padj")))
    
    # Adjusts column names
    setnames(ex2, c("GeneID", "Symbol", "Description"), c("ENTREZID", "SYMBOL", "GENENAME"))
    
    # Gets raw data
    GeneID <- as.integer(rownames(exraw))
    exraw <- as.data.table(exraw)
    
    # Updates column names
    exraw$ENTREZID <- GeneID
    
    # Merges FC and raw data together
    mer <- merge(ex2, exraw, by = "ENTREZID")
    return(mer)
  }
}

```

With this function now defined, a data set from GEO can now be downloaded using the code below. This code downloaded a GEO RNAseq dataset that evaluated the differences in gene expression levels of the nucleus accumbens between cigarette smokers and non-smokers [1] and was saved as `data`. 

```{r 2, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Downloads smokers vs non-smokers data set
data <- GEODataDownload(DS = "GSE171936",
                              gpl = "GPL21290",
                              gsm = "00000000110100000000100001000000000110001010110111010000000001010001000001110010001010000100011100000000000000001110100110000000011000000001000000000010000000000001000000000001010000000001100000000000001001X1X11100010000000",
                              PlateAnnotInfo = "Only needed if the data was generated using MicroArrays",
                              GenerateMetaData = "Only needed if the data was generated using MicroArrays",
                              Technology = "RNAseq")

# Confirms successful download of dataset by showing first three rows of data set
data.table(head(as.data.frame(data),3), filter = 'top', options = list(pageLength = 10, scrollX = TRUE, scrollY = "400px", autoWidth = TRUE))

```

The metadata for this data set was downloaded from the GEO website and saved to my working directory. This metadata csv file was loaded into this session using the code below.

```{r 3, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Loads meta data csv file into session
DT <- fread("metadata.csv")

# Displays first few lines of meta data csv file
head(DT)

```


# Exploratory Analysis
With data set and meta data loaded into the session, exploratory analysis can begin using the code below. First we must evaluate the data set for any missing information and adjust accordingly.

```{r 4, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Isolates sample name, current smoking status, and age at death from meta data 
meta <- DT[, c("GEO_Accession (exp)", "current_smoking_status", "age_at_death")]
meta <- unique(meta)

# Identifies meta data for control samples
controls <- meta[meta$current_smoking_status == "control"]
head(controls)
dim(controls)

# Identifies meta data for test samples
cases <- meta[meta$current_smoking_status == "case"]
head(cases)
dim(cases)

# Identifies samples with unknown smoking status
unknown.status <- meta[meta$current_smoking_status == "indeterminant"]
unknown.status
dim(unknown.status)

# Identifies samples with unknown age
unknown.age <- meta[is.na(meta$age_at_death),]
unknown.age
dim(unknown.age)

# Isolates sample names with incomplete information
unknowns <- unknown.status$`GEO_Accession (exp)`

# Identifies columns containing data from samples with incomplete information
unknown_1 <- which(colnames(data) == unknowns[1])
unknown_2 <- which(colnames(data) == unknowns[2])
unknown_1
unknown_2

# Removes samples with unknown smoking status
data.filtered <- data[, -c(213,215)]

# Confirms samples with unknown smoking status are removed
head(data[, c(212:216)])
head(data.filtered[, c(212:216)])

# Removes samples with unknown smoking status from meta data
meta <- meta[meta$current_smoking_status != "indeterminant"]

```

Here we see our data contains 171 control samples and 50 test samples. We also see that two samples, GSM5237811 and GSM5237813, have an unknown smoking status and an unknown age. Since we are interested in smoking status, we can remove these two samples. As shown in the code above, we identified the columns these samples occupy in our `data` and remove them, saving this new data set as `data.filtered`. We also update the meta data information to remove these two samples we are not interested in. 

## Age Comparison Between Groups
With our data filtered, we can compare the ages between smokers and non-smokers 

```{r 5, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Finds average age at death for controls and test samples
mean(controls$age_at_death)
mean(cases$age_at_death)
median(controls$age_at_death)
median(cases$age_at_death)

# Plots ages differences between controls and test samples
boxplot(controls$age_at_death, cases$age_at_death,
        main = "Controls vs Tests Ages at Death",
        names = c("Controls", "Tests"),
        col = c("orange", "darkgreen"))

# Isolates ages and smoking status 
ages <- meta$age_at_death
status <- meta$current_smoking_status

# Evaluates difference in ages between groups
t.test(ages ~ status, data = meta)

```

Here we see that the controls in this data set have a slightly lower average age at death, ~45.8 years old, compared to the test samples, ~47.9 years old. However, the median age of the controls is slightly higher, ~48.8 years old, compared to the test samples, ~47.16 years old. Performing a t-test on these samples confirms that there is no statistically significant difference in the age at death between smokers and non-smokers in this data set. 

## Normalization of Expression Data
Now let's explore the expression data. We can first evaluate if normalization is necessary for our data set. 

```{r 6, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Isolates all expression data
all_expr_data <- data.filtered[, -c(1:6)]

# Isolates expression data for controls
control_expr_data <- subset(all_expr_data, select = controls$`GEO_Accession (exp)`)

# Isolates expression data for test samples
case_expr_data <- subset(all_expr_data, select = cases$`GEO_Accession (exp)`)

# Converts isolated expression data to a true numeric matrix
expr_matrix <- as.matrix(all_expr_data)

# Sets the row names as the  ENTREZID
rownames(expr_matrix) <- data.filtered$ENTREZID  

# Ensures meta dataframe rows align with the columns of expr_matrix
meta_df <- as.data.frame(meta)
rownames(meta_df) <- meta_df$`GEO_Accession (exp)`
meta_df <- meta_df[colnames(expr_matrix), ] 

# Reconstructs Log2 transformations for evaluation of data
logged_counts <- log2(expr_matrix + 1)

# Evaluates if expression data needs normalization
boxplot(logged_counts, main="Sample Distributions Prior to Normalization", 
        las=2, outline=FALSE, col="skyblue", xlab="Samples", ylab="Log2 Raw Counts")


# Converts isolated data to a strict numeric matrix for normalization
count_matrix <- as.matrix(all_expr_data)
rownames(count_matrix) <- data.filtered$ENTREZID

# Ensures meta dataframe rows align with the columns of count_matrix
meta_df <- as.data.frame(meta)
rownames(meta_df) <- meta_df$`GEO_Accession (exp)`
meta_df <- meta_df[colnames(count_matrix), ]

# Performs structural RNA-seq normalization (VST)
dds <- DESeqDataSetFromMatrix(countData = count_matrix, 
                              colData = meta_df, 
                              design = ~ current_smoking_status)
dds <- estimateSizeFactors(dds)
vsd <- varianceStabilizingTransformation(dds, blind = FALSE)
norm_counts <- assay(vsd) 

# Evaluates if normalization of data was appropriate
boxplot(norm_counts, main="Sample Distributions of Normalized Data", 
        las=2, outline=FALSE, col="skyblue", xlab="Samples", ylab="Log2 Norm Counts")

```

Here we see that the expression data does appear to be experiencing batch effects or technical variations prior to normalization. Once the data is normalized using VST normalization, our sample distributions no longer appear to be influenced by any potential batch effects! 

## Principle Component Analysis
With the data now normalized, we can perform PCA on the normalized data set to determine if specific data attributes result in any clustering. 

```{r 7, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Calculates gene variance 
rv <- apply(norm_counts, 1, var)

# Performs PCA on all genes in normalized data set
rv <- order(rv, decreasing = TRUE)
pca_res <- prcomp(t(norm_counts[rv, ]))

# Builds PCA data frame using metadata and first two principle components
pca_df <- data.frame(
  PC1 = pca_res$x[,1],
  PC2 = pca_res$x[,2],
  Smoking_Status = as.factor(meta_df$current_smoking_status),
  Age_at_Death   = as.numeric(meta_df$age_at_death)
)

# Checks variance in smoking status
ggplot(pca_df, aes(x=PC1, y=PC2, color=Smoking_Status)) + 
  geom_point(size=4, alpha=0.8) + theme_minimal() + 
  labs(title="PCA: Colored by Smoking Status", x="PC1", y="PC2")

# Checks variance in age at death
ggplot(pca_df, aes(x=PC1, y=PC2, color=Age_at_Death)) + 
  geom_point(size=4, alpha=0.8) + scale_color_viridis_c() + theme_minimal() + 
  labs(title="PCA: Colored by Age at Death", x="PC1", y="PC2")

# Identifies genes contributing most to variance
loadings <- abs(pca_res$rotation[, 1])
top_features <- sort(loadings, decreasing = TRUE)[1:10]
top_features
ENTREZIDs <- names(top_features)
gene_names <- as.data.frame(data.filtered[, c("ENTREZID", "SYMBOL", "GENENAME"), with = FALSE])
matched_symbols <- gene_names$SYMBOL[match(as.character(ENTREZIDs), as.character(gene_names$ENTREZID))]
matched_details <- gene_names$GENENAME[match(as.character(ENTREZIDs), as.character(gene_names$ENTREZID))]
print(matched_symbols)

```

Using variance from all genes in the data set for the PCA plot above, there appears to be no clustering based on smoking status or age with our normalized data. Additionally, when evaluating which genes contribute the most to variance, RPS29 is found to contribute the most variance in this data set. Additionally, the top 10 genes that contribute the most variance are:

- RPS29      
- XIST        
- PWAR1       
- P2RY12      
- LOC107985363
- LOC112268323
- MUC6        
- CX3CR1      
- RPS4Y1 
- LOC112268125

# Differential Expression Analysis
Now differential expression analysis can be performed on our data!

```{r 8, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Prepares data for differential expression analysis
meta_clean <- na.omit(meta_df)
counts_clean <- count_matrix[, rownames(meta_clean)]
counts_clean <- round(counts_clean)
mode(counts_clean) <- "integer"

# Forces smoking status to be a factor 
meta_clean$current_smoking_status <- factor(meta_clean$current_smoking_status, 
                                            levels = c("control", "case"))

# Re-initializes DESeq dataset structure
dds_analysis <- DESeqDataSetFromMatrix(countData = counts_clean,
                                       colData = meta_clean,
                                       design = ~ current_smoking_status)

# Performs differential expression analysis based on smoking status
dds_analysis <- DESeq(dds_analysis) 

# Extracts contrast results table based on significance
res <- results(dds_analysis, 
               contrast = c("current_smoking_status", "case", "control"),
               alpha = 0.05, 
               pAdjustMethod = "fdr")

# Corrects for bias in low count genes
target_coef <- grep("current_smoking_status_case_vs_control", resultsNames(dds_analysis), value = TRUE)
res_shrunk  <- lfcShrink(dds_analysis, 
                         coef = target_coef, 
                         type = "normal") 
# Generate overall evaluation summary metrics
summary(res_shrunk)

```

In the summary table we see that, of the 25,274 genes in this data set, approximately 60 of them were differentially expressed between smokers and non-smokers. Looking closer, 6 of these genes were up-regulated while 54 were down-regulated. Since these are just genes that were either up-regulated or down-regulated, they should be filtered down to only look at significantly differentially expressed genes.

```{r 9, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Filter out the statistically significant genes
sig_genes_df <- as.data.frame(res_shrunk)
sig_genes_df$ENTREZID <- rownames(sig_genes_df)

# Merge with gene annotations
annotations <- as.data.frame(data.filtered[, c("ENTREZID", "SYMBOL", "GENENAME"), with = FALSE])
sig_genes_annotated <- merge(sig_genes_df, annotations, by = "ENTREZID")

# Filter for significance: FDR < 0.05 
sig_final_table <- subset(sig_genes_annotated, padj < 0.05)

# Print top 10 most significant differentially expressed genes
head(sig_final_table[order(sig_final_table$padj), ], 10)

```

Focusing on just the statistically significant genes with an adjusted p-value < 0.05, we see that only 6 total genes meet this threshold:

- glucosidase II alpha subunit (GANAB)
- SEC61 translocon subunit alpha 1 (SEC61A1)
- tRNA-Cys (TRNC)
- tRNA-Tyr (TRNY)
- surfeit 4 (SURF4)
- methionyl-tRNA synthetase 1 (MARS1)

Both TRNC and TRNY are up-regulated with log2 Fold Change of 0.1765037 and 0.1807380, respectively. They each have an adjusted p-value of ~0.025. The other four genes are down-regulated. GANAB has a log2 Fold change of -0.1400501, SEC61A1 is -0.1229403, SURF4 is -0.1263831, and MARS1 is -0.1177700. These down-regulated genes all had an adjusted p-value of ~0.025 except for MARS1 which had an adjusted p-value of ~0.039. 

Let's check the fit of this model using the code below:

```{r Q10, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Evaluates mathematical model fit quality
plotDispEsts(dds_analysis, 
             main = "Model Evaluation: Dispersion Estimates")

```

This diagnostic plot does not perfectly align with the model's predicted trajectory, however, this is to be expected with human models. Considering the data still follows the general path of the fitted line with no other unexpected shapes or trajectories in the plot, we can infer that the model works well with our data. 

A volcano plot would provide excellent visualization to show our significant genes identified earlier. 

```{r 11, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}

# Prepares data for volcano plot
volcano_data <- as.data.frame(res_shrunk)
volcano_data$ENTREZID <- rownames(volcano_data)
annotations_df <- as.data.frame(data.filtered[, c("ENTREZID", "SYMBOL"), with = FALSE])
volcano_data <- merge(volcano_data, annotations_df, by = "ENTREZID", all.x = TRUE)

# Sets threshold for significant genes
volcano_data$Significance <- "Not Significant"
volcano_data$Significance[volcano_data$padj < 0.05 & volcano_data$log2FoldChange > 0] <- "Upregulated in Smokers"
volcano_data$Significance[volcano_data$padj < 0.05 & volcano_data$log2FoldChange < 0] <- "Downregulated in Smokers"
volcano_data$SYMBOL[is.na(volcano_data$SYMBOL) | volcano_data$SYMBOL == ""] <- volcano_data$ENTREZID[is.na(volcano_data$SYMBOL) | volcano_data$SYMBOL == ""]
sig_labels <- subset(volcano_data, padj < 0.05)

# Generates Volcano Plot
ggplot(volcano_data, aes(x = log2FoldChange, y = -log10(pvalue))) +
  
  # Sets visualization for non-significant genes
  geom_point(data = subset(volcano_data, Significance == "Not Significant"), 
             aes(color = Significance), alpha = 0.4, size = 1.5) +
  
  # Sets visualization for significant genes
  geom_point(data = subset(volcano_data, Significance != "Not Significant"), 
             aes(color = Significance), alpha = 1.0, size = 4) +
  
  # Sets colors based on significance
  scale_color_manual(values = c("Not Significant" = "grey75", 
                                "Upregulated in Smokers" = "firebrick2", 
                                "Downregulated in Smokers" = "dodgerblue3")) +
  geom_hline(yintercept = -log10(max(sig_labels$pvalue, na.rm=TRUE)), linetype = "dashed", col = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", col = "grey40") +
  theme_minimal() +
  labs(title = "Volcano Plot: Smokers vs Non-Smokers Profile",
       x = "Log2 Fold Change (Effect Size)",
       y = "-Log10 Raw P-Value") +
  theme(legend.position = "bottom",
        plot.title = element_text(face = "bold", size = 14),
        legend.title = element_blank())

```

In this volcano plot, we see that most genes in this data set are significantly differentially expressed. However, this plots shows 6 total points, representing genes, that are significant. We see 4 genes (GANAB, SEC61A1, SURF4, and MARS1) that are down-regulated in smokers and 2 genes (TRNC and TRNY) that are up-regulated in smokers. This suggests that addressing these 6 genes and better understanding their pathways may be a route for therapeutic interventions to address relapse risk in patients attempting to quit smoking.


# References
[1] Markunas CA, Semick SA, Quach BC, Tao R et al. Genome-wide DNA methylation differences in nucleus accumbens of smokers vs. nonsmokers. Neuropsychopharmacology 2021 Feb;46(3):554-560. PMID: 32731254